A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-31 paper.tex
Clarify narrative in several places.

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Despite its status as ``the ideal data structure for
population genomic analysis''~\citep{rasmussen2014genome}, there has been
little discussion of its computational properties. We discuss the
relationship between the classical concept of an ARG and the
recently introduced succinct tree sequence data structure.
\end{enumerate}

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the 
stochastic process generating genealogies in the presence of 
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent}, 
as well as, more commonly, the concrete realisations of the 
ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes:
we need to know from which parent the node inherited
the segments of genome to the left and right of the breakpoint.
The simplest way to break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG (the most basic downstream
operation for an ARG data structure).
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

A gARG is a graph in which nodes represent monoploid genomes (i.e.,
haploids have one genome, diploids two, etc.) and edges represent
genetic inheritance between an ancestor and a descendant.
Each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

There are some interesting and important consequences to this focus on genomes
rather than events.
Firstly, a gARG may contain nodes that are both samples \emph{and}
internal nodes in the local trees.
This ability is useful when we have pedigree information
along with genetic data
\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
where we may have many generations of internal sample nodes
whose genomes have been sequenced.
Another situation in which it is important that
we do not assume samples are all contemporary ``leaf'' nodes
(or that all internal nodes are either coalescences or recombinations)
is when we are incorporating ancient genomes into ARG
inference~\citep{speidel2021inferring,wohns2022unified}.
Similarly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective.
Secondly, because we do not classify nodes by event type
(common ancestor or recombination) there is no limitation
on the complexity what can be modelled---multiple recombinations
and coalescences can happen simultaneously in the same genome.
In large deeply sequenced pedigrees such complexities are commonplace.
Finally, there is a subtle and important point about how recombination
is represented: which genome in the organismal lifecycle
do nodes represent?
In principle, we can intepret nodes as representing genomes
at any point in the life cycle [supplementary Figure ?]. Hudson
views nodes as representing gametes~\citep{hudson1983properties}.
However, we argue that the most intuitive
approach is to let nodes represent the genomes present in diploid
individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. [TODO finish up here with more
clarity on what this means in terms of the required specifity
of recombination events, and put in a forward ref to the simplification
section where we talk about stacking up recombination events
and their identifiability.]

@@

\section*{Simplification}\label{ARG_simplification}
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy,
retaining the ``meaningful'' topology within local trees
and maintaining the identity of nodes across different trees.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the highlighted tree (covering positions 60 to 70)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The best
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

@@

The main reasons for wanting precise estimates of all recombination
events are to either compute a likelihood of the inferred ARG
under a statistic model, or to study the recombinants in
detail~\citet{rasmussen2022espalier}.
% Explain that we're really interested in the details of recombination
% itself with a small number of samples.
In the latter case, we might
have a small number of samples~\citep{guo2022recombination}.

% If we want to use the coalescent, then really the assumptions
% need to make sense.
While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$. This
assumption ensures that multiple events (i.e., simultaneous
coalescence and recombination) don't occur at the same
time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

Another assumption of the coalescent with recombination which prevents complex
combinations of events occurring simultaneously, is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
% Clearly multiple recombinations (and other complex events) will
% occur in the ancestry of such datasets, and we should not be
% limited in what we can infer by the data structure that we
% use to represent the results.

>>>>

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Despite its status as ``the ideal data structure for
population genomic analysis''~\citep{rasmussen2014genome}, there has been
little discussion of its computational properties. We discuss the
relationship between the classical concept of an ARG and the
recently introduced succinct tree sequence data structure.
\end{enumerate}

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process generating genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisations of the
ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes:
we need to know from which parent the node inherited
the segments of genome to the left and right of the breakpoint.
The simplest way to break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event older than a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

A gARG is a graph in which nodes represent monoploid genomes (i.e.,
haploids have one genome, diploids two, etc.) and edges represent
genetic inheritance between an ancestor and a descendant.
Each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

% This paragraph is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?
There are some interesting and important consequences to this focus on genomes
rather than events.
Firstly, a gARG may contain nodes that are both samples \emph{and}
internal nodes in the local trees.
This ability is useful when we have pedigree information
along with genetic data
\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
where we may have many generations of internal sample nodes
whose genomes have been sequenced.
Another situation in which it is important that
we do not assume samples are all contemporary ``leaf'' nodes
(or that all internal nodes are either coalescences or recombinations)
is when we are incorporating ancient genomes into ARG
inference~\citep{speidel2021inferring,wohns2022unified}.
Similarly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective.
Secondly, because we do not classify nodes by event type
(common ancestor or recombination) there is no limitation
on the complexity that can be modelled---multiple recombinations
and coalescences can happen simultaneously in the same genome,
a common occurance in large deeply sequenced
pedigrees~\citep{hayes20191000,RosFreixedes2022}.
Finally, there is a subtle and important point about how recombination
is represented: which genome in the organismal lifecycle
do nodes represent?
In principle, we can interpret nodes as representing genomes
at any point in the life cycle [supplementary Figure ?]. Hudson
views nodes as representing gametes~\citep{hudson1983properties}.
However, we argue that the most intuitive
approach is to let nodes represent the genomes present in diploid
individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. [TODO finish up here with more
clarity on what this means in terms of the required specifity
of recombination events, and put in a forward ref to the simplification
section where we talk about stacking up recombination events
and their identifiability.]

@@

\section*{Simplification}\label{ARG_simplification}
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy,
retaining the ``meaningful'' topology within local trees
and maintaining the identity of nodes across different trees.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a tree node that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the highlighted tree (covering positions 60 to 70)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

@@

The main reasons for wanting precise estimates of all recombination
events are to either compute a likelihood of the inferred ARG
under a statistical model, or to study the recombinants in
detail~\citet{rasmussen2022espalier}.
% Explain that we're really interested in the details of recombination
% itself with a small number of samples.
In the latter case, we might
have a small number of samples~\citep{guo2022recombination}.

Beyond cases in which visual inspection of the trees and the
effects of recombination is feasible, the main application
of having completely precise estimates of recombination events is
to evaluate the likelihood of the estimate under some
statistical model. The only statistical model available
is the coalescent with recombination
and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
In this case, the utility of a ``full`` ARG for some dataset
is limited by how well the dataset fits the assumptions
of the model.

% If we want to use the coalescent, then really the assumptions
% need to make sense.
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$.
% This
% assumption ensures that multiple events (i.e., simultaneous
% coalescence and recombination) don't occur at the same
% time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Another core assumption of the coalescent model is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
% Clearly multiple recombinations (and other complex events) will
% occur in the ancestry of such datasets, and we should not be
% limited in what we can infer by the data structure that we
% use to represent the results.

While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting}, [something
something]
